Gradual Soundness: Lessons from Static Python

### 2022-04-26 pj-7.1/response.txt

TODO
- write a quick thank-you and summary of changes
- add section + page numbers to the responses

> Review #45A
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> The paper makes a contribution to gradual typing in theory and practice. The
> subject language is Static Python, which adds a compile-time type discipline to
> Python.
>
> On the theory side, the paper describes a formal model created with an
> explorative tool for semantics modeling and prototyping. This model includes a
> static semantics and a dynamic semantics. The dynamic semantics is tested
> against a regression test suite.
>
> On the practical side, the paper describes the design of Static Python as well
> as experiments with an implementation of Static Python. This implementation
> allegedly improves the throughput of a closed source implementation of a web
> server by 3.6%. It improves the run time of several microbenchmarks
> significantly.
>
> Assessment and comments (for authors)
> -------------------------------------
> ## strengths
>
> * The main strength of the paper is clearly the favorable evaluation of a
>   particular datapoint in the design space of gradual typing.
> * The formal model enabled the discovery of problems in a industrial strength
>   implementation of Static Python, at least in its type checker.
>
> ## weaknesses
>
> * the most significant evaluation is on a closed source program
> * the microbenchmarks potentially gain more efficiency from progressive
>   primitive types as most of them are computationally intensive.
> * the theoretical evaluation is only probabilistic with unclear probability
>
> ## comments
>
> I like the paper, it is generally well readable. There are a few places, where
> I'd expect more guidance from the authors, rather than expecting the reader to
> interpolate.
>
> ### Table 1
>
> The caption is misleading. The data is the cost of creating and using
> shallow/concrete dictionary types in typed and untyped contexts.  I was
> initially confused by the 0 entries. I'd prefer you put --- to indicate that no
> casts is needed.  You should indicate that $N$ is the length of the dictionary.

Thanks, we've updated the caption, explained N, and used --- for no cast.

> ### enforcement of CheckedDict[K,V]
>
> *... gets enforced in constant-time* What if V is also a CheckedDict?

We've added an explanation of nested types (they do run in constant time).

> ### p8, 2
>
> *Nom [...] flexible point in the design space*
> Explain this point in two sentences.

Done.

> ### p8, se 2.5
>
> *...static semantics is progressive...*
> It seems like the second paragraph *Primitive types...* explains the
> progressive aspect. You should affirm that in the text.

We've expanded the first paragraph and revised the second paragraph. The first
paragraph focuses on the "normal" progressive aspects. The second focuses on
gradual typing.

> ### p9, -3
>
> *one token*
> Is this a variable size token or fixed size (if V is CheckedDict? see above)

It is always a fixed-size pointer.

> ### p10, 2
>
> *only if*
> Is that an `iff`?

No, the "only if" is intentional. An "iff" would be correct, but the paragraph
in which this sentence appears is about one direction in particular.

> ### p10, -17
>
> *Redex's random testing ... to check type soundness*
>
> That seems a bit shaky. What's the baseline / ground truth / oracle for these tests?
> In the end, all that testing only gives you a probabilistic guarantee.
> Can you give an estimate of this probability? (I mean, quantitative)

These are property-based tests. Our model defines a soundness property for terms
and asks Redex to generate random terms in the hope of finding a counterexample.
So, the soundness property is the ground truth; these random tests are all
checking soundness.

One could imagine a probabilistic guarantee about what kinds of terms the tests
explore, but the guarantee would depend on two complicated parameters: our
model's grammar and Redex's algorithm for exploring grammars (Klein & Findler,
SFP 2009). These parameters make it difficult to formulate a guarantee. On top
of that, **explaining** the guarantee to readers would be a huge challenge.

Property-based testing is shaky compared to formal verification, but it's still
a helpful tool for finding bugs.  Indeed, we found several bugs in our model
thanks to these property-based tests (none of which had implications for Static
Python itself).

> ### p10, -2
>
> *All but one...*
>
> Which soundness bug was **not** fixed and why?

All soundness bugs have been fixed at this time.

> ### p11, -3
>
> `Dyn` should be `Any` in Python.

The revised paper explains why we use the name `Dyn`.

> ### p12, fig 3, 4
>
> * `str` should be italic
> * why is `object` part of the expression language?

Fixed `str` and removed `object`.

(`object` was there as a simple instance of the `Object` type. Since it's not
needed to explain type boundary issues, we can remove it.)

> Line 6: `C.f(expr)`
> * looks like a static method call to me, is that right? That's usually trivial
>   compared with the dynamic dispatch of a real method call.
> * how do you invoke a method on an object?

Oh dear. This was not meant to be a static method call, but we were far too
aggressive when simplifying the model.

The revision uses `expr.f(expr)` instead and looks up the class type of the
receiver to type check the call.

> * how do you create an object? Is there a syntax like `C(expr)`?

Syntax for making an object was missing from the paper (not from SP or Redex!).
We've added it, thanks!

> ### fig 4
>
> Some evaluation types don't make sense
>
> * can Optional types be nested? E.g. Optional[Optional[int]]   doesn't make
>   sense to me
> * what about Optional[None] ??
> * or Optional[Dyn]??

Yes, those types can all be written. Just like surface unions (section 4.1), we
assume that Optional evaluation types get normalized in a standard way. The
normal forms for these examples are: `Optional[int]`, `None`, and `Dyn`.

> BTW, I find it confusing that surface types are `T`, but evaluation types are
> `S`, given that surface starts with s.

Good point, we've switched `S` and `T`.

> ### table 2
>
> Misleading caption: this is about the cast from Any to an evaluation type.

We have reworded the caption.

> ### p14, fig 5
>
> * rule `M-App` seems too limiting. Literally read, it's not possible to invoke
>   inherited methods because the method `f0` must be defined in class `C0` in
>   which it is defined. This rule affirms the class-method conjecture from
>   above.

We had envisioned the lookup `Env(C0)` as searching both `C0` and its parents
for all visible fields and methods.

The revised figure writes `Env*(C0)` instead and includes a note to this effect.
(The revision also uses `\Gamma` instead of `Env`)

> * the `object` rule for subtyping contradicts Table 2, which says that *objects
>   accepts any value expect primitives*. So `S0` must not be a primitive; but
>   primitives are not part of the model. This is not consistent!

We've moved the Table 2 comment about primitives to a footnote (because, as you
rightly point out, they are not part of the model).

> * given transitivity, the last rule could be simplified by omitting the
>   assumption `C2 <: C1` and changing the conclusion to `C0 <: C1`.

Thanks!

> ### p15, -16
>
> *better left to the Redex mechanization*
>
> It would be useful to show the excerpt for method calls here, as they seem to
> be nonstandard.

Method calls in Redex are standard; they work like calls in (Static) Python.
The confusion here is due to our over-simplified model in the submission, which
gave the impression that only a class can receive a method call.

> ### p24
>
> references [23] and [24] and [41] have broken DOIs

Fixed.

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45B
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> The contribution of the paper is the following:
>
> - The paper describes Static Python, a project developed at Instagram to add
>   types to Python code in a gradual manner, with the main objective of
>   improving performance. The project has already shown its effectiveness. The
>   authors (informally) describe both the language (the gradual type system) and
>   how this is implemented to achieve better performance.
>
> - The authors have developed an implementation in PLT redex of a subset of
>   Static Python, focusing on features related to gradual typing. Using this
>   implementation, they did an extended testing to show:
>   1) that this implementation is sound
>   2) that this implementation gives the same results of Static Python.
>
> Assessment and comments (for authors)
> -------------------------------------
> Assessment: The contributions of the paper summarized above are certainly
> useful for the PL community, so I support acceptance. The empirical evaluation
> seems to me (but I am definitely not an expert in such kind of contributions)
> well-driven and adequate. I have some reservations on the formalization,
> explained in the comments below, which should be solved/improved in the final
> version.
>
> Detailed comments:
>
> Page 3 Mechanization: you should explain better what do you mean by
> "soundness". I thought of standard soundness in the sense that "well-typed
> programs do not go stuck", but here you say "property-based soundness tests",
> what do you mean more precisely?

Section 4 in the revised paper gives more details.

In short, we started with the property "well-typed programs do not get stuck"
and tested it on thousands of terms generated by Redex.

> Section 4.1 line 4: In the formal syntax in Figure 3, Object seems not to be an
> instance of C, so class C(C) seems not to include the case C = Object.
> In the "Other expressions" you mention object fields reads and writes, but in
> the formal syntax in Figure 3 you have C.x and C.x = expr, I was expecting
> expr.x and expr.x = expr' instead.

We've made several changes in response:

- `Object` is no longer part of the grammar
- The grammar comes with a note that the set `C` must contain `Object`
- Field reads and writes use `expr` instead of `C` (example: C.x is now expr.x)

> Figure 3: it would be better to use a more abstract style with metavariables as
> it is customary now, e.g.
> prog = stmt_1, ..., stmt_n
> expr = ... [x_1:expr_1, ..., x_n:expr_n] | expr[expr'] |

We appreciate the suggestion, but prefer to keep the current grammar because it has
no substripts and uses a uniform style for the "sequence" terminals `prog` and `Env`.

> Page 13 line -7: A dynamically-typed variable -> A dynamically-typed function

No, "variable" is there on purpose because we don't know statically what a
dynamically typed variable points to. It might not be a function.

> Figure 5 rule D-App: in Figure 3 a function f seems not to be allowed to have
> type Dyn in the enviroment

This is allowed, but only indirectly. The definition of a function `f_0` can't
directly say that `f_0` has type Dyn, but it can declare the type `Dyn -> Dyn`
and then apply `f_0(f_0)` to assign the dynamic type to a function.

(Note that `x` and `f` are undifferentiated variable names. The model uses `x`
 in some definitions and `f` in others as hints to readers, but it's fine to
 name a function `x_2` and so on.)

> Also, Object seems to be the top type, hence also Bool and Int are seen as
> classes, right?

Yes, Bool and Int are technically classes. The paper says little about this,
but the Redex model (and Static Python) give them full support as classes.

> In rules D-Set and CD-Set, the type None seems to be used as a "Unit" type. The
> fact that these two types are identified seems strange.

Yes it is a bit strange, but it helps keep the model small for the paper.
(Really, the best choice would be to move these rules into a typing judgment
for **statements** rather than expressions. That's what the Redex model does.)

> In subtyping rules, which is the relation between two Optional types?

Fixed, thank you for catching this omission.

> Why do you use metavariable Env_0 rather than just Env in the rules?

We've removed the subscript.

> page 15 line 4: I would not say "upcast" since there is no explicit cast here.

Ok. We have removed this explanatory sentence.

> My main criticism about your formalization of the typing rules is that, in my
> opinion, they should include generation of code annotated with casts.
> Moreover,it should be made clear that they are algorithmic. I mean, what I
> expect is that there is an algorithm which, taken an expression expr and an
> environment Env, produces a type (or a typing error) and the corresponding
> annotated code, inserting casts. I am sure that this is what happens in your
> PLT Redex implementation, and I think that this should be reported.

The typing judgment in Redex is algorithmic, but the simple judgment in the
paper is not. Section 4.2.2 in the revised paper makes this clear.

> Section 4.3: I would perhaps eliminate this section. As it is, it says almost
> nothing.

We disagree. The section is a proof sketch that ill-typed values cannot "sneak
in" through the dynamic type.

> Section 6.3 I miss a conclusion/summary of the analysis here.

We've added a conclusion section.

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45C
> ===========================================================================
>
> Paper summary (for authors)
> ---------------------------
> This paper presents the design of Static Python, which introduces
> gradual typing for Python.  It is already in use by engineers at
> Instagram.  One of the notable features in the design is that a
> programmer can have some control over the choice of what kind of
> invariant can be enfored by types: untyped (no guarantee), shallow (in
> the sense that types guarantee only the top-level type constructor),
> and concrete (where full type information is relevant).  The use of
> concrete types promotes code optimization, which can be also an
> incentive to programmers.
>
> The paper presents an overview of the language and its run-time
> system, followed by the description of the formal type system.  The
> type system is formalized with PLT Redex.  Although no formal
> metatheory is developed, the model has been checked by Redex's random
> testing tools and test suites.  Subtleties in the language features
> that are not modeled but implemented are discussed, too.  The section
> on evaluation reports that 3.7% imprevement in CPU efficiency.
>
> Assessment and comments (for authors)
> -------------------------------------
> The language design mixing concrete and shallow types is interesting.
> It seems that the language has been accepted by engineers and used for
> a large software project.  Nevertheless I have a few reservations to
> accept the paper in its current form.
>
> The paper puts emphasis on "gradual soundness" at the beginning but,
> after reading the paper, I failed to see what it really was.  The
> phrase is never mentioned on page 3 and afterwards.  This is
> unfortunate because this is the phrase used even in the title!

We've updated the paper to mention "gradual soundness" in the following
sections: 2.3, 2.5, 6.3, and 8.

> It's not clear how faithfully the formal model presented in Section 4
> models the implementation.  In particular, C-Sub allows the type of an
> expression can be changed to Dyn (using the consistent subtyping) and
> Matr allows conversion from Dyn to any non-Dyn type.  So, the type
> system seems to give any expression (that references only declared
> variables, functions, classes) any type.  However, apparently, some
> expressions are rejected by the implementation.  I have no idea what's
> really going on.  (I tried to access the Redex model but it wasn't
> accessible.)

We've added a disclaimer near the top of section 4.2.2 to say that the
paper presents declarative typing rules whereas the Redex model implements a
typing algorithm.

> The distinction between surface types and evaluation types is
> interesting but its implication is not so clear.  Are type arguments
> to Dict completely ignored by the type system?  If so, what is the
> point of writing type arguments in a program?  I may misunderstand
> something and am wondering if my question is related to the paragraph at
> the bottom of p.11, "we omit the surface typing judgment, which is a
> kind of linter that merely scans typed code for logical errors".  What
> are the logical errors mentioned here?  If type arguments to Dict are
> completely ignored, an alternative design might have been like this:
> instead of providing two classes Dict and CheckedDict, a programmer
> could use the raw type Dict and fully parameterized types Dict[S0,S1].

The discussion is intentionally vague because Static Python does not have firm
guidelines for what the surface type checker (linter) should do. For Dict,
it currently ignores the type arguments so the only reasons to write them are
for documentation and for other type checkers (Pyre, mypy) to check them.
In the future, though, Static Python may do more surface checks.

Using the raw type Dict and parameterized type Dict[S0, S1] seems problematic
because these types describe different data structures. The types look similar,
but a value with type Dict[S0, S1] cannot be used where a Dict is expected.

> The related work section is light.  It cites relevant papers and
> describes what they discuss but doesn't really compare with the
> present work in depth.

We have expanded the related work.

> In summary, I'd recommend that the paper be revised by:
> * clarifying gradual soundness,
> * clarifying the relationship between the model and the implementation,
> * clarifying the role of surface types,
> * enhancing the discussion on related work.
> It would be appreciated if you make the code of the Redex model accessible.

We apologize that the Redex model was inaccessible. For convenience, here is
the link:

  <https://github.com/brownplt/insta-model>

> Detailed comments:
>
> Table 1: Why does a constructor call for CheckedDict pay the cost in
> O(N)? (What does N stand for?)

N is the number of keys and values in the table (we have updated the table
caption to explain). The constructor needs to validate each element.

> * It seems that efficient run-time checking depends on the fact that
>   the set of type instantiations CheckedDict[S0, S1] is finite and
>   known at compile-time.  How is it achieved?  (Is it because there
>   are no user-defined generic classes?)  My guess above may be wrong.
>   In that case, it's not very clear how tags are implemented.

Section 2.3 and 3 in the revision say more about the "how." Static Python
implements these tags as pointers into a run-time table of all generic
instantiations that have appeared.

> p.12: "Evaluation types S are a subset of the surface types".
> Is "Dict", which is an evaluation type, a surface type?

No, Dict is not a surface type in the model. We have rephrased "subset" to
avoid this confusion.

### 2022-04-26 pj-7.1/paper.tex
pj: first-draft response + updated paper

%% TODO 2022-04-08 from Pyre team
%% Hi Carl! The small notational question I mentioned about in the paper was in
%% Figure 5 -- there's this "M-APP" rule that describes how classmethod invocation
%% is handled. I was trying to figure out how inherited classmethod would work
%% with this rule. Say I have a class B that defines a method f0(B, S0) -> S1, and
%% then I have another class D that inherits B which does not define additional
%% method. Now given Env0 \vdash expr0: S0, how can Env0 \vdash D.f0(expr0): S1 be
%% proven? If I understand correctly, the rule only looks up the type of f0 in
%% Env0(D), i.e. the body of the derived class. But the real definition of f0 is
%% syntactically located in the base class B. Does the notation system used here
%% assume that every function definition in base class gets automatically copied
%% into the derived class? Or should there be an additional preconditions added to
%% the "M-APP" rule saying that f0 is allowed to exist in any of the superclass of
%% D?

%% 2022-04-24 More model issues:
%% - C.f(e) uses a class type, not an instance ... how to invoke method on instance?!
%% ... we should really apologize for a too-simple model
%% - Object is not clearly in C
%% - introduce c again as variable name? gee why not expr.method()?

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{S}}
\newcommand{\spteval}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptstr}{\typefont{Str}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptflout}{\typefont{Flout}}
\newcommand{\sptdyn}{\typefont{Dyn}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptset}[1]{\paramtype{\typefont{Set}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{Dict}}
\newcommand{\sptrawchkdict}{\typefont{CheckedDict}} %% not a real SP type, but useful for TeX
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\sptrawchkdict}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\sptclass}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\wtsub}{\vdash}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spconsist}{\spcompat}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spstr}{\codefont{str}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spcnew}[2]{\spapp{#1}{#2}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is to provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a tag check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a
    check for concrete dicts that have exactly the same key and value
    types.
    To implement such checks efficiently, \SP{} stores a run-time registry
    of instantiated concrete types.
    Every concrete dict carries a pointer into this registry to enable
    constant-time type validation, even for nested concrete types.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options keep run-time costs low and let programmers decide between
concrete and shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

For performance-critical applications, \SP{} includes a set of primitive types
that describe booleans and sized numbers.
Example types include \code{int64}, \code{uint64}, \code{double}.
There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
The purpose of these types is to allow unboxed values and arithmetic at
runtime.
Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
refinement over the semantics of basic Python types:
they permit fewer behaviors statically to enable the unboxing.
Operations such as \code{and} (boolean conjunction), for example, do not accept
a mix of primitive and non-primitive arguments.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness on thousands of examples (1,600 expressions and 11,000
programs).
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

For most of the \numSPtest{} tests, the translation is automatic.
A few tests required hand-pruning to remove features that the model does not
handle (\totalnum{52}).
\SP{} has \numSPdiff{} other tests (\totalnum{\numSPtotal{}}) that we did not
use because they fall outside the scope of the model.
\Cref{s:impl} summarizes the out-of-scope features; \cref{a:banned-test} gives
a detailed categorization.

@@

The rest of this section illustrates our Redex model using
a small formalization.
\Cref{s:surface} presents a surface syntax to highlight the boundaries between
typed and untyped code.
\Cref{s:eval-types} explains how types get enforced at run-time.
\Cref{s:ts} argues that the overall approach toward boundaries is sound.

@@

@@

\begin{figure}[t]
  \begin{langarray}
    \spteval & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptstr \langmid
      \sptobject \langmid
      \sptclass \langmid
    \\ & &
      \sptrawpydict \langmid
      \sptchkdict{\spteval}{\spteval} \langmid
      \sptoptional{\spteval}
  \end{langarray}

@@

@@

>>>>

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{S}}
\newcommand{\spteval}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptstr}{\typefont{Str}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptflout}{\typefont{Flout}}
\newcommand{\sptdyn}{\typefont{Dyn}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptset}[1]{\paramtype{\typefont{Set}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{Dict}}
\newcommand{\sptrawchkdict}{\typefont{CheckedDict}} %% not a real SP type, but useful for TeX
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\sptrawchkdict}{#1, #2}}
\newcommand{\sptenv}{\Gamma} %{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\sptclass}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\wtsub}{\vdash}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spconsist}{\spcompat}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spstr}{\codefont{str}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spcnew}[2]{\spapp{#1}{#2}}
\newcommand{\spenvapp}[2]{\spapp{#1{}^*}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is to provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a tag check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a
    check for concrete dicts that have exactly the same key and value
    types.
    To implement such checks efficiently, \SP{} stores a run-time registry
    of instantiated concrete types.
    Every concrete dict carries a pointer into this registry to enable
    constant-time type validation, even for nested concrete types.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options keep run-time costs low and let programmers choose between
concrete and shallow types.
This choice is one aspect of gradual soundness in \SP{}.

Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

For performance-critical applications, \SP{} includes a set of primitive types
that describe booleans and sized numbers.
Example types include \code{int64}, \code{uint64}, \code{double}.
There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
The purpose of these types is to allow unboxed values and arithmetic at
runtime.
Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
refinement over the semantics of basic Python types:
they permit fewer behaviors statically to enable the unboxing.
Operations such as \code{and} (boolean conjunction), for example, do not accept
a mix of primitive and non-primitive arguments.
These primitive types are another main aspect of gradual soundness.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness.
Specifically, we asked Redex to generate well-formed terms (using its default
algorithm) and checked whether the well-typed terms that terminated within a
fixed time budget (1,200 expressions and 11,000 programs in a typical run)
produced only well-typed values and allowed errors.
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
We translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

For most of the \numSPtest{} tests, the translation is automatic.
A few tests required hand-pruning to remove features that the model does not
handle (\totalnum{52}).
\SP{} has \numSPdiff{} other tests (\totalnum{\numSPtotal{}}) that we did not
use because they fall outside the scope of the model.
\Cref{s:impl} summarizes the out-of-scope features and \cref{a:banned-test} gives
a detailed categorization.

@@

%% TODO gaps between paper and redex
%% (if these are small enough, explain here ... maybe with minor notes further down
%%  if bigger, consider a "threats" or "gaps" section)
%% - typing is not algorithmic
%% - few python features (well that's obvious)
%% - few expressions (very subtle to introduce here/now)

The rest of this section illustrates our Redex model using a small
formalization.
%% TODO EXAMPLE simplification
\Cref{s:surface} presents a surface syntax to highlight the boundaries between
typed and untyped code.
\Cref{s:eval-types} explains how types get enforced at run-time.
\Cref{s:ts} argues that the overall approach toward boundaries is sound.

@@

  \medskip
  \mbox{Convention: the set of all class types $\sptclass$ contains the top type $\sptobject$}

@@

\begin{figure}[t]
  \begin{langarray}
    \spteval & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptstr \langmid
      \sptclass \langmid
    \\ & &
      \sptrawpydict \langmid
      \sptchkdict{\spteval}{\spteval} \langmid
      \sptoptional{\spteval}
  \end{langarray}

@@

By contrast to our Redex model, the typing judgment in~\cref{f:eval-type} is
not syntax directed.
The rules do not describe a typechecking algorithm.
Instead these rules declare various conclusions and their preconditions to
illustrate how \SP{} reasons about types---dynamic or otherwise.

@@

\subsubsection{Conclusion}
